Hypercoercions and a Framework for Equivalence of Cast Calculi

### 2019-10-22 paper.tex
minor changes
progress

\newcommand{\TOOdyn}[0]{\star}
\newcommand{\TOOpre}[1]{#1}
\newcommand{\POOunit}[0]{\mathtt{Unit}}
\newcommand{\POOfun}[2]{#1 \shortrightarrow #2}
\newcommand{\POOprod}[2]{#1 \times #2}
\newcommand{\POOsum}[2]{#1 \plus #2}
\newcommand{\eOOvar}[1]{#1}
\newcommand{\eOOsole}[0]{\mathtt{unit}}
\newcommand{\eOOlam}[4]{\lambda^{#1\rightarrow{}#2}#3.#4}
\newcommand{\eOOapp}[2]{#1 \; #2}
\newcommand{\eOOcons}[2]{\mathtt{cons} \; #1 \; #2}
\newcommand{\eOOcar}[1]{\mathtt{fst} \; #1}
\newcommand{\eOOcdr}[1]{\mathtt{snd} \; #1}
\newcommand{\eOOinl}[1]{\mathtt{inl} \; #1}
\newcommand{\eOOinr}[1]{\mathtt{inr} \; #1}
\newcommand{\eOOcase}[3]{\mathtt{case} \; #1 \; #2 \; #3}
\newcommand{\eOOcast}[4]{#1 \langle \cOOcast{#2}{#3}{#4} \rangle}
\newcommand{\eOOblame}[1]{\mathtt{blame} \; #1}
\newcommand{\cOOcast}[3]{#1 \Rightarrow^{#2} #3}
\newcommand{\oOOinj}{\mathtt{dyn}}
\newcommand{\oOOsole}{\mathtt{unit}}
\newcommand{\oOOfun}{\mathtt{fun}}
\newcommand{\oOOcons}{\mathtt{cons}}
\newcommand{\oOOinl}{\mathtt{inl}}
\newcommand{\oOOinr}{\mathtt{inr}}
\newcommand{\oOOblame}[1]{\mathtt{blame} \; #1}
\newcommand{\vOOcast}[2]{#1\langle#2\rangle}
\newcommand{\vOOfun}[3]{\mathtt{fun} \; #1 \; #2 \; #3}
\newcommand{\vOOtt}[0]{\mathtt{unit}}
\newcommand{\vOOcons}[2]{\mathtt{cons}\;#1\;#2}
\newcommand{\vOOinl}[1]{\mathtt{inl}\;#1}
\newcommand{\vOOinr}[1]{\mathtt{inr}\;#1}
%\newcommand{\rOOsucc}[1]{#1}
%\newcommand{\rOOfail}[1]{#1}
\newcommand{\rOOsucc}[1]{\mathtt{succ}\;#1}
\newcommand{\rOOfail}[1]{\mathtt{fail}\;#1}

\newcommand{\kOOmt}[0]{\mathtt{stop}}
\newcommand{\kOOconsI}[3]{[\mathtt{cons} \; \square \; \langle#1,#2\rangle ]#3}
\newcommand{\kOOconsII}[2]{[\mathtt{cons} \; #1 \; \square]#2}
\newcommand{\kOOinl}[1]{[\mathtt{inl} \; \square]#1}
\newcommand{\kOOinr}[1]{[\mathtt{inr} \; \square]#1}
\newcommand{\kOOappI}[3]{
	[\mathtt{app} \; \square \; \langle#1,#2\rangle ]#3
}
\newcommand{\kOOappII}[2]{
	[\mathtt{app} \; #1 \; \square]#2}
\newcommand{\kOOcar}[1]{[\mathtt{fst} \; \square]#1}
\newcommand{\kOOcdr}[1]{[\mathtt{snd} \; \square]#1}
\newcommand{\kOOcaseI}[4]{
	[\mathtt{case} \; \square \; \langle#1,#3\rangle \; \langle#2,#3\rangle ]#4}
\newcommand{\kOOcaseII}[4]{
	[\mathtt{case} \; #1 \; \square \; \langle#2,#3\rangle ]#4}
\newcommand{\kOOcaseIII}[3]{
	[\mathtt{case} \; #1 \; #2 \; \square]#3}
\newcommand{\kOOcast}[2]{
	[\square \langle #1 \rangle]#2}

@@

To summarize, the primary contributions of this paper are:
\begin{itemize}
\item hypercoercions, a new space-efficient cast representation, which
  has a structurally recursive composition and a more compact memory
  representation.
\item a framework in Agda for proving the correctness of cast
  represantations.
\item a formal proof that hypercoercion respects the semantics
  of the \lazyD{} cast calculus.
\end{itemize}

@@

In this section we review the coercions in normal form of
\citet{siek2012interpretations} to motivate the design of
hypercoercions.  We ignore sum types and product types in this
section, because \citet{siek2012interpretations} did not discuss
them. We assume a basic familiarity with coercions and direct readers
who are not familiar with them to \citet{siek2012interpretations}.
%We 
%will highlight the similarity of normal coercion and hypercoercion, and hope 
%this would convince you that hypercoercion should be as difficult to implement 
%as normal coercions.
%\todo[inline]{I don't think we should try to convince the reader that
%  hypercoercions are easier to implement. We don't have real evidence
%  for that.}

\begin{figure}
	\[
	\begin{array}{rclr}
	\stxrule{I}{injectable types (\lazyD)}{
		\POOfun{T}{T}}
	\stxrule{I}{injectable types (\lazyUD)}{
		\POOfun{\TOOdyn}{\TOOdyn}
	}
	\stxrule{c}{coercions}{
		\ncInj{I} \mid
		\ncProj{I}{l} \mid
		\ncId \mid
		\ncFail{l} \mid
		\ncSeq{c}{c} \mid
		\ncFun{c}{c}
	}
	\stxrule{\bar{c}}{wrapper coercions}{	
		\ncInj{I} \mid
		\ncFun{\hat{c}}{\hat{c}} \mid
		\ncSeq{\ncFun{\hat{c}}{\hat{c}}}{\ncInj{I}}
	}
%	\stxrulecont{
%		\ncProd{\hat{c}}{\hat{c}} \mid
%		\ncSeq{\ncProd{\hat{c}}{\hat{c}}}{\ncInj{I}} \mid
%		\ncSum{\hat{c}}{\hat{c}} \mid
%		\ncSeq{\ncSum{\hat{c}}{\hat{c}}}{\ncInj{I}} \mid	
%	}
	\stxrule{\hat{c}}{normal coercions}{
		\bar{c} \mid
		\ncId \mid
		\ncFail{l} \mid
		\ncProj{I}{l} \mid
		\ncSeq{\ncProj{I}{l}}{\ncFail{l}} \mid
		\ncSeq{\ncProj{I}{l}}{\ncInj{I}}
	}
	\stxrulecont{
		\ncSeq{\ncProj{I}{l}}{\ncFun{\hat{c}}{\hat{c}}} \mid
		\ncSeq{\ncProj{I}{l}}{\ncSeq{\ncFun{\hat{c}}{\hat{c}}}{\ncInj{I}}}
	}
	\end{array}
	\]
	\caption{Syntax of Normal Coercion}
	\label{fig:normal-coercion}
\end{figure}

\todo[inline]{Give a summary statement about the shape of the normal form. 
-JS
\\---\\The three-part shape is not obvious before the transformation at 
the end of the next paragraph. So the summary is given after that. -KC}

\figref{fig:normal-coercion} defines the normal coercion.
Types, pre-types, and blame labels are as before.
Let $I$ range over injectable types. An injectable type is a type that can 
be cast directly to (injection) and from (projection) $\TOOdyn$. The 
definition of injectable type depends on blame strategy. With \lazyD, all 
pre-types are injectable. With \lazyUD, only $\POOfun{\TOOdyn}{\TOOdyn}$
is injectable. 
Let $c$ range over coercions, a coercion is either injection, projection, 
identity, failure, sequencing, or function coercion.
Let $\bar{c}$ range over wrapper coercions, which represent casts in $
\vOOcast{v}{c}$.
Let $\hat{c}$ range over normal coercions. If we inline $\bar{c}$ and 
re-order the cases, the definition of $\hat{c}$ becomes:
\[
\begin{array}{rclr}
\stxrule{\hat{c}}{normal coercions}{
	\ncFun{\hat{c}}{\hat{c}} \mid
	\ncSeq{\ncProj{I}{l}}{\ncFun{\hat{c}}{\hat{c}}} \mid
	\ncSeq{\ncFun{\hat{c}}{\hat{c}}}{\ncInj{I}} \mid
	\ncSeq{\ncProj{I}{l}}{\ncSeq{\ncFun{\hat{c}}{\hat{c}}}{\ncInj{I}}} \mid
	\ncSeq{\ncProj{I}{l}}{\ncFail{l}}
}
\stxrulecont{
	\ncId \mid
	\ncProj{I}{l} \mid
	\ncInj{I} \mid
	\ncSeq{\ncProj{I}{l}}{\ncInj{I}} \mid
	\ncFail{l} \mid
}
\end{array}
\]

@@

\section{Definition of Hypercoercion} \label{sec:hypercoercion-definition}

@@

	\caption{Definition of hypercoercion (HC)}
	\label{fig:hypercoercion}
\end{figure}

This section presents our first contribution, the definition of hypercoercion. 
The design of hypercoercion is motivated by observations on coercion normal 
forms: a normal coercion has at most three parts; projections are always at the 
beginning; injections and failures are always at the end. Hypercoercions have 
similar shape: a hypercion either is $\hyperCoercionI$, the identity cast for 
$\TOOdyn$, or contains three parts, where the first part is either a 
projection or a no-op while last part is either an injection, a failure, or a 
no-op.

@@

\subsection{\lazyD{} hypercoercion}
\label{sec:ld-hc}

@@

@@

\funrule{
	\continue{v_1}{\kOOappI{e_2}{\rho}{\kappa}}}{
	\sOOinspect{e_2}{\rho}{
		wrap(\kOOappII{v_1}{\kappa})
	}}{}
\hifunrule{
	\continue{v_2}{(\kOOappII{(\hcvOOfun{c_1}{\rho}{x}{e}{c_2})}{\kappa})}}{
\begin{cases}
	\sOOinspect{e}{\rho[x:=v']}{ext(c_2,\kappa)} \\
\;\;\;\;\;\sidecond{applyCast(v_2,c_1) = \rOOsucc{v'}}
\\
\sOOhalt{l} \\
\;\;\;\;\;\sidecond{applyCast(v_2,c_1) = \rOOfail{l}}
\end{cases}
}{}

@@

>>>>

\newcommand{\TOOdyn}[0]{\star}
\newcommand{\TOOpre}[1]{#1}
\newcommand{\POOunit}[0]{\mathtt{Unit}}
\newcommand{\POOfun}[2]{#1 \shortrightarrow #2}
\newcommand{\POOprod}[2]{#1 \times #2}
\newcommand{\POOsum}[2]{#1 \plus #2}
\newcommand{\eOOvar}[1]{#1}
\newcommand{\eOOsole}[0]{\mathtt{unit}}
\newcommand{\eOOlam}[4]{\lambda^{#1\rightarrow{}#2}#3.#4}
\newcommand{\eOOapp}[2]{(#1 \; #2)}
\newcommand{\eOOcons}[2]{\mathtt{cons} \; #1 \; #2}
\newcommand{\eOOcar}[1]{\mathtt{fst} \; #1}
\newcommand{\eOOcdr}[1]{\mathtt{snd} \; #1}
%\newcommand{\eOOinl}[1]{\mathtt{inl} \; #1}
%\newcommand{\eOOinr}[1]{\mathtt{inr} \; #1}
\newcommand{\eOOinl}[1]{\mathtt{inl} \; #1}
\newcommand{\eOOinr}[1]{\mathtt{inr} \; #1}
\newcommand{\eOOcase}[3]{\mathtt{case} \; #1 \; #2 \; #3}
\newcommand{\eOOcast}[4]{#1 \langle \cOOcast{#2}{#3}{#4} \rangle}
\newcommand{\eOOblame}[1]{\mathtt{blame} \; #1}
\newcommand{\cOOcast}[3]{#1 \Rightarrow^{#2} #3}
\newcommand{\oOOinj}{\mathtt{dyn}}
\newcommand{\oOOsole}{\mathtt{unit}}
\newcommand{\oOOfun}{\mathtt{fun}}
\newcommand{\oOOcons}{\mathtt{cons}}
\newcommand{\oOOinl}{\mathtt{inl}}
\newcommand{\oOOinr}{\mathtt{inr}}
\newcommand{\oOOblame}[1]{\mathtt{blame} \; #1}
\newcommand{\vOOcast}[2]{#1\langle#2\rangle}
\newcommand{\vOOfun}[3]{\mathtt{fun} \; #1 \; #2 \; #3}
\newcommand{\vOOtt}[0]{\mathtt{unit}}
\newcommand{\vOOcons}[2]{\mathtt{cons}\;#1\;#2}
\newcommand{\vOOinl}[1]{\mathtt{inl}\;#1}
\newcommand{\vOOinr}[1]{\mathtt{inr}\;#1}
%\newcommand{\rOOsucc}[1]{#1}
%\newcommand{\rOOfail}[1]{#1}
\newcommand{\rOOsucc}[1]{\mathtt{succ}\;#1}
\newcommand{\rOOfail}[1]{\mathtt{fail}\;#1}

\newcommand{\kOOmt}[0]{\mathtt{stop}}
\newcommand{\kOOconsI}[3]{[\mathtt{cons} \; \square \; \langle#1,#2\rangle ]#3}
\newcommand{\kOOconsII}[2]{[\mathtt{cons} \; #1 \; \square]#2}
\newcommand{\kOOinl}[1]{[\mathtt{inl}\; \square]#1}
\newcommand{\kOOinr}[1]{[\mathtt{inr}\; \square]#1}
\newcommand{\kOOappI}[3]{
	[\square \; \langle#1,#2\rangle]#3
}
\newcommand{\kOOappII}[2]{
	[#1 \; \square]#2}
\newcommand{\kOOcar}[1]{[\mathtt{fst} \; \square]#1}
\newcommand{\kOOcdr}[1]{[\mathtt{snd} \; \square]#1}
\newcommand{\kOOcaseI}[4]{
	[\mathtt{case} \; \square \; \langle#1,#3\rangle \; \langle#2,#3\rangle ]#4}
\newcommand{\kOOcaseII}[4]{
	[\mathtt{case} \; #1 \; \square \; \langle#2,#3\rangle ]#4}
\newcommand{\kOOcaseIII}[3]{
	[\mathtt{case} \; #1 \; #2 \; \square]#3}
\newcommand{\kOOcast}[2]{
	[\square \langle #1 \rangle]#2}

@@

To summarize, the primary contributions of this paper are:
\begin{itemize}
\item hypercoercions, a new space-efficient cast representation, which
  has a structurally recursive composition and a more compact memory
  representation.
\item a framework in Agda for proving the correctness of \lazyD{} cast
  represantations.
\item a formal proof that hypercoercions respect the semantics
  of the \lazyD{} cast calculus.
\end{itemize}

@@

In this section we review the
coercions~\citep{henglein1994dynamic,herman2010space} and the normal
form of \citet{siek2012interpretations} to motivate the design of
hypercoercions.  We ignore sum types and product types in this
section, because \citet{siek2012interpretations} did not discuss
them. We assume a basic familiarity with coercions and direct readers
who are not familiar with them to \citet{siek2012interpretations}.

\begin{figure}
	\[
	\begin{array}{rclr}
	\stxrule{I}{injectable types (\lazyD)}{
		\POOfun{T}{T} \mid \POOunit}
	\stxrule{I}{injectable types (\lazyUD)}{
		\POOfun{\TOOdyn}{\TOOdyn} \mid \POOunit
	}
	\stxrule{c}{coercions}{
		\ncInj{I} \mid
		\ncProj{I}{l} \mid
		\ncId \mid
		\ncFail{l} \mid
		\ncSeq{c}{c} \mid
		\ncFun{c}{c}
	}
	\stxrule{\bar{c}}{wrapper coercions}{	
		\ncInj{I} \mid
		\ncFun{\hat{c}}{\hat{c}} \mid
		\ncSeq{\ncFun{\hat{c}}{\hat{c}}}{\ncInj{I}}
	}
%	\stxrulecont{
%		\ncProd{\hat{c}}{\hat{c}} \mid
%		\ncSeq{\ncProd{\hat{c}}{\hat{c}}}{\ncInj{I}} \mid
%		\ncSum{\hat{c}}{\hat{c}} \mid
%		\ncSeq{\ncSum{\hat{c}}{\hat{c}}}{\ncInj{I}} \mid	
%	}
	\stxrule{\hat{c}}{normal coercions}{
		\bar{c} \mid
		\ncId \mid
		\ncFail{l} \mid
		\ncProj{I}{l} \mid
		\ncSeq{\ncProj{I}{l}}{\ncFail{l}} \mid
		\ncSeq{\ncProj{I}{l}}{\ncInj{I}}
	}
	\stxrulecont{
		\ncSeq{\ncProj{I}{l}}{\ncFun{\hat{c}}{\hat{c}}} \mid
		\ncSeq{\ncProj{I}{l}}{\ncSeq{\ncFun{\hat{c}}{\hat{c}}}{\ncInj{I}}}
	}
	\end{array}
	\]
	\caption{Syntax of coercions and normal forms}
	\label{fig:normal-coercion}
\end{figure}

\figref{fig:normal-coercion} gives the grammar for coercions, written
$c$ and also coercions in normal form, written $\hat{c}$.
%
To review, an injection $I!$ takes a value from an injectiable type
$I$ to type $\TOOdyn$. An injectable type is simply a type that can be
cast directly to (injection) and from (projection) $\TOOdyn$. The
definition of injectable type depends on blame strategy. With \lazyD,
all pre-types are injectable. With \lazyUD, only
$\POOfun{\TOOdyn}{\TOOdyn}$ is injectable.
%
A projection $I?^l$ takes a value from type $\TOOdyn$ to type $I$, or
halts the program and blames $l$ if the value is of a different type.
The coercion $\iota$ is the identity, $\bot^{l}$ is the coercion that
always fails and blames $l$, and $\ncSeq{c_1}{c_2}$ applies $c_1$ and
then $c_2$ in sequence. The function coercion $\ncFun{c_1}{c_2}$
applies $c_1$ to the argument of a function and $c_2$ to the return
value.

Coercions come with a reduction relation so it is natural to ask what
are their normal forms. The most important of the reduction rules are
the ones that handle when an injection is followed by a projection,
which for \lazyUD{} are:
\begin{align*}
  I! ; I?^l &\longmapsto \iota \\
  I_1! ; I_2?^l & \longmapsto \bot^l & \text{if } I_1 \neq I_2
\end{align*}
and the rule for pushing sequences below function coercions:
\begin{align*}
  (c_1 \to c_2) ; (c_3 \to c_4) & \longmapsto
  (c_3 ; c_1) \to (c_2 ; c_4)
\end{align*}
Of course there are also rules that remove identity coercions.

\todo[inline]{UNDER CONSTRUCTION -Jeremy}

Let $\bar{c}$ range over wrapper coercions, which represent casts in $
\vOOcast{v}{c}$.  Let $\hat{c}$ range over normal coercions. If we
inline $\bar{c}$ and re-order the cases, the definition of $\hat{c}$
becomes:
\[
\begin{array}{rclr}
\stxrule{\hat{c}}{normal coercions}{
	\ncFun{\hat{c}}{\hat{c}} \mid
	\ncSeq{\ncProj{I}{l}}{\ncFun{\hat{c}}{\hat{c}}} \mid
	\ncSeq{\ncFun{\hat{c}}{\hat{c}}}{\ncInj{I}} \mid
	\ncSeq{\ncProj{I}{l}}{\ncSeq{\ncFun{\hat{c}}{\hat{c}}}{\ncInj{I}}} \mid
	\ncSeq{\ncProj{I}{l}}{\ncFail{l}}
}
\stxrulecont{
	\ncId \mid
	\ncProj{I}{l} \mid
	\ncInj{I} \mid
	\ncSeq{\ncProj{I}{l}}{\ncInj{I}} \mid
	\ncFail{l} \mid
}
\end{array}
\]

@@

\section{Definition of Hypercoercions} \label{sec:hypercoercion-definition}

@@

	\caption{Definition of hypercoercions (HC)}
	\label{fig:hypercoercion}
\end{figure}

This section presents our first contribution, the definition of
hypercoercions.  The design of hypercoercions is motivated by
observations on coercion normal forms: a normal coercion has at most
three parts; projections are always at the beginning; injections and
failures are always at the end. Hypercoercions have similar shape: a
hypercion either is $\hyperCoercionI$, the identity cast for
$\TOOdyn$, or contains three parts, where the first part is either a
projection or a no-op while last part is either an injection, a
failure, or a no-op.

@@

\subsection{\lazyD{} hypercoercions}
\label{sec:ld-hc}

@@

\subsection{Compact Memory Representation for Hypercoercions}

\todo[inline]{UNDER CONSTRUCTION (a couple paragraphs) -Jeremy}

@@

\funrule{
	\continue{v_1}{\kOOappI{e_2}{\rho}{\kappa}}}{
	\sOOinspect{e_2}{\rho}{
		wrap(\kOOappII{v_1}{\kappa})
	}}{}

\hifunrule{
	\continue{v_2}{(\kOOappII{(\hcvOOfun{c_1}{\rho}{x}{e}{c_2})}{\kappa})}}{
\begin{cases}
\sOOinspect{e}{\rho[x:=v']}{ext(c_2,\kappa)} \\
\sidecond{applyCast(v_2,c_1) = \rOOsucc{v'}}
\\
\sOOhalt{l} \\
\sidecond{applyCast(v_2,c_1) = \rOOfail{l}}
\end{cases}
}{}

@@

\todo[inline]{Explain that this is an unusual formulation of a weak
  bisimulation.  We promise it will all be OK regarding how it is used
  in the correctness proof for S(C). -JS}
